Low Complexity Adaptive Beamformer\\ for Active Sonar Imaging

### 2015-05-22 LCAforActiveSonarImaging/lca.tex
la2pdf commit - 2015:05:22	16:43:47	(Fri22.May)

Due to the limitations imposed on hardware, a lot of research is put into improving the part of the digital image formation process. A core component here is the beamformer, a spatial band-pass filter used on the wave field to estimate each pixel value. To allow the use of wide-band signals a coarse steering is first applied by delaying the sensor channels, followed by an individual weighting of them. Using complex weights, this implies fine adjustment of both the amplitude and the phase of the spatial filter. While conventional beamformers apply static weight functions, adaptive methods dynamically tailor the weights to the wave field.

One of the best studied adaptive methods is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}. It

 solves an optimization problem to come up with a weight set that yields unity gain in the desired direction, while minimizing the effect of noise and interference from other directions. The result is often good looking non-linear image with improved contrast and detail resolution. However, MVDR is neither statistically nor numerically robust by default, and typically needs to average data in space and time, and regularization. This constrains the solution space of the weights.

We will show that when applied to experimental data, the weights typically chosen by a robustified MVDR yield responses that are typically symmetric and steered within a fraction of the -3dB resolution of the respective window. Then we make Kaiser windows of different types, and with varying degree of steering, that combined form a discrete version of the MVDR solution space.

@@

@@

By definition, the beamformer output $z[n]$ can now be expressed as the weighted sum of all the delayed data samples:
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
\end{align}
where $w_m$ is the weight factor assigned to channel $m$. With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer. A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.

Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield. In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality. The MVDR beamformer is one such method. It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}. This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{weights}
\end{gather}
where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. Since we pre-steer our data to every pixel in the image we simplify (\ref{weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering. To estimate $\R$ we compute a sample covariance matrix $\eR$. In this computation we perform some degree of:
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
\end{itemize}%
%
% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009a}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}. 
%
Combined, these steps will also ensure a numerically well conditioned $\eR$.

>>>>

Due to the limitations imposed on hardware, a lot of research is put into improving the part of the digital image formation process. A core component here is the beamformer, a spatial band-pass filter used on the wave field to estimate each pixel value. A way to do this while allowing wide-band signals is to steer coarsely by applying delays to the sensor channels, followed by an individual weighting of them. Using complex weights, this implies fine adjustment of both the amplitude and the phase of the spatial filter. While conventional beamformers apply static weight functions, adaptive methods dynamically tailor the weights to the wave field.

One of the best studied adaptive methods is the minimum variance distortion-less response (MVDR) beamformer~\cite{Capon1969}. It computes a weight set that minimizes the power of noise and interference solves an optimization problem to come up with a weight set that yields unity gain in the desired direction, while minimizing the effect of noise and interference from other directions. The result is often good looking non-linear image with improved contrast and detail resolution. However, MVDR is neither statistically nor numerically robust by default, and typically needs to average data in space and time, and regularization. This constrains the solution space of the weights.

We will show that when applied to experimental data, the weights chosen by a robustified MVDR yield responses that tend to be symmetric and steered within a fraction of the -3dB resolution of the respective window. Then we create Kaiser windows of different types, and with varying degree of steering, that combined form a discrete version of the MVDR solution space.

@@

This article is outlined as follows: In Section \ref{methods}, we introduce the concept of adaptive beamforming and provide details on the MVDR and LCA method. Then, in \ref{maptogpu}, we investigate the complexity issue, discuss means for reducing arithmetic complexity, and detail an implementation that makes efficient use of the GPU's parallel resources. The final design is assessed in Sections \ref{images_and_benchmarks} and \ref{discussion}, where we provide benchmarks, comparisons with similar CPU implementations, and measures of how efficiently our implementation makes use of the GPU's resources.

% - How much oversampling
% - How many parameters, how great a span.
% 
% - Experiemental data, Holmengraa/cross
% 
% - Note on efficiency.

@@

By definition, the beamformer output $z[n]$ can be expressed as the weighted sum of all the delayed data samples:
\begin{align}
z[n] = \w\H[n]\x[n] = \bmat{w_0[n]\\w_1[n]\\\vdots\\w_{M-1}[n]}^H \bmat{x_0[n]\\x_1[n]\\\vdots\\x_{M-1}[n]},\label{z}
\end{align}
where $w_m$ is the weight factor assigned to channel $m$. With static weights this would be referred to as the conventional delay-and-sum (DAS) beamformer. A large variety of weighting functions exists here for trading lateral resolution for improved noise suppression (contrast), but one always ends up with a compromise between the two~\cite{Harris1978}.

Various adaptive beamformers target this limitation by allowing the weights to change for each pixel to better fit the dynamic nature of the incoming wavefield. In other words, they attempt to use the \emph{a priori} information present in the data to improve image quality. The MVDR beamformer is one such method. It finds the set of complex weights that minimizes the beamformer's expected output power, while ensuring unity gain in the look direction~\cite{Capon1969}.

\begin{align}
\underset{\w[n]}{\argmin}\, E\{|z[n]|^2\} &= \underset{\w[n]}{\argmin}\, \w[n]\x\x\H\w\H[n]\\
\text{subject to } \w[n]\1 &= 1
\end{align}

This is a convex optimization problem that can be solved using Lagrange multipliers to yield the solution
\begin{gather}
\vec w[n] = \frac{\Ri[n]\a}{\a\T\Ri[n]\a},\label{weights}
\end{gather}
where $\a$ is a steering vector and $\R=E\{\x[n]\x\H[n]\}\in\mathbb{C}^{M,M}$ is the spatial covariance matrix for the full array. Since we pre-steer our data to every pixel in the image we simplify (\ref{weights}) by substituting $\a$ with a row vector $\1$ that represents broadside phase-steering. To estimate $\R$ we compute a sample covariance matrix $\eR$. In this computation we perform some degree of:
\begin{itemize}
\item \emph{spatial averaging} to avoid signal cancellation by decorrelating coherent echoes~\cite{Kailath1985};
\item \emph{temporal averaging} over an interval comparable to the pulse length (one to five samples) to maintain true speckle statistics~\cite{Synnevag2009a};
\item \emph{diagonal loading} to improve robustness to parameter errors~\cite{Cox1987,Maksym1979}.
\end{itemize}%
%
% will perform some degree of \emph{spatial averaging} to avoid signal cancellation caused by coherent echoes~\cite{Kailath1985}, \emph{temporal averaging} to maintain true speckle statistics~\cite{Synnevag2009a}, and \emph{diagonal loading} to improve robustness to parameter errors ~\cite{Cox1987,Maksym1979}. 
%
Combined, these steps will also ensure a numerically well conditioned $\eR$.
